---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE,
  message = FALSE, 
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(faintr)
library(brms)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_bw() + theme(plot.background = element_blank()))

custom_palette <- c("#009E73", "#B22222", "#0072B2", "#D55E00")

scale_colour_discrete <- function(...) {
  scale_colour_manual(..., values = custom_palette)
}
scale_fill_discrete <- function(...) {
   scale_fill_manual(..., values = custom_palette)
}
```

# faintr

<!-- badges: start -->
[![R-CMD-check](https://github.com/michael-franke/faintr/workflows/R-CMD-check/badge.svg)](https://github.com/michael-franke/faintr/actions)
[![Codecov test coverage](https://codecov.io/gh/michael-franke/faintr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/michael-franke/faintr?branch=main)
<!-- badges: end -->

The **faintr** (FActorINTerpreteR) package provides convenience functions for 
interpreting [brms](https://github.com/paul-buerkner/brms) model fits for data 
from factorial designs. It allows for the extraction and comparison of posterior 
draws for a given design cell, irrespective of the encoding scheme used in the model.

Currently, **faintr** provides the following functions:

* `get_cell_definitions` returns information on the predictor variables and how 
they are encoded in the model.
* `extract_cell_draws` returns posterior draws for one subset of design cells.
* `compare_groups` returns summary statistics of comparing two subsets of design cells.

## Installation

You can install the development version from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("michael-franke/faintr")
```

## Examples

In this section, we shortly introduce how to use the package. For a more detailed 
overview, please refer to the vignette.

We will use a preprocessed version of R's built-in Titanic data set:

```{r data, echo=FALSE}
data <- Titanic %>% as_tibble() %>% rename(Count = n)
data_agg <- data %>% group_by(Class, Sex, Age) %>% summarize(Total = sum(Count))

data <- full_join(data %>% filter(Survived == "Yes") %>% select(-Survived), 
                  data_agg)

data
```
The data set contains the following variables:

* `Class`: Passenger class
* `Sex`: Sex of the passenger
* `Age`: Age group of the passenger
* `Count`: Count of passengers who survived
* `Total`: Total number of passengers

Below, we regress the counts of passengers who survived as a function of their class, 
sex, and age group using a Binomial logistic regression model fitted with `brms`:

```{r model-fitting, results='hide'}
fit <- brm(Count | trials(Total) ~ Class + Sex + Age,  
           data = data, 
           family = binomial(link = "logit"),
           seed = 123)
```

```{r model-summary}
summary(fit)
```
To obtain information on the factors and the encoding scheme used in the model, 
we can use `get_cell_definitions`:

```{r cell-defs}
get_cell_definitions(fit)
```

The output reveals that we used dummy coding for factors `Class`, `Sex` and 
`Age`, where `1st`, `Female`, and `Adult` are the reference levels, respectively.

To obtain posterior draws for a specific design cell, we can use `extract_cell_draws`. 
For instance, draws for women in the second class can be extracted like so:

```{r cell-draws}
extract_cell_draws(fit, Sex == "Female" & Class == "2nd" & Age == "Adult") %>% 
  head()
```
Parameter `colname` allows changing the default column name in the output, which 
facilitates post-processing of cell draws, e.g., for plotting or summary statistics. 
Here, we extract the draws for each level of `Class` (averaged over `Sex` and `Age`) 
and visualize the results:

```{r plot, out.width="70%"}
draws_1st  <- extract_cell_draws(fit, Class == "1st", colname = "1st")
draws_2nd  <- extract_cell_draws(fit, Class == "2nd", colname = "2nd")
draws_3rd  <- extract_cell_draws(fit, Class == "3rd", colname = "3rd")
draws_crew <- extract_cell_draws(fit, Class == "Crew", colname = "Crew")

draws_class <- tibble(draws_1st, draws_2nd, draws_3rd, draws_crew) %>% 
  pivot_longer(cols = everything(), names_to = "class", values_to = "draws")

draws_class %>% 
  ggplot(aes(x = draws, color = class, fill = class)) +
  geom_density(alpha = 0.4)
```

Finally, we can compare two subsets of design cells with `compare_groups`. Here,
we compare the odds of surviving between female passengers in the first class and
male passengers in all but the first class:

```{r group-comp}
compare_groups(fit, 
               Sex == "Female" & Class == "1st",
               Sex == "Male" & Class != "1st"
               )
```
If one of two group specifications is left out, we compare against the grand mean:

```{r group-comp_grand-mean}
compare_groups(fit, 
               Class == "Crew"
               )
```

